Co-citations in context

### 2019-06-11 cocit_r.tex
safety branch chacko:
continuing edits

%% Documentclass:
%% Network Neuroscience
%\documentclass[finalfonts,NETN]{stjour}
\documentclass[NETN]{stjour}

@@

%%%%%%%%%%% Article Set-Up %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Article Type:
%% Default is Research.

@@

Citation and network analysis of  scientific literature reveals information on semantic relationships between publications, collaboration between scientists, and the practice of citation  and TESTING THE MERGE process itself~\citep{garfield_citation_1955,de_solla_price_networks_1965,newman_structure_2001,Shi:2010:CHI:1816123.1816131,patience_pmid28560354}. Co-citation, the frequency with which two documents are cited together in other documents provides additional insights, including the identification of semantically related documents, fields, specializations, and new ideas in science~\citep{small_co-citation_1973, marshakova-shaikevich_co-citation_1973,boyack_co-citation_2010, 10.3389/frma.2018.00020}.

\citep{uzzi_atypical_2013} used a novel approach for co-citation analysis, using 17.9 million articles and their cited references from the Web of Science (WoS) as input, to characterize a subset of highly cited articles with respect to both novel and conventional combinations of prior research. The frequency with which references were co-cited was calculated and expressed as journal pair frequencies (observed co-citation frequencies). Expected values were generated from Monte Carlo simulations under a random graph model. Observed frequencies were then normalized (shifted and scaled) to averaged expected values from ten simulations and termed as \emph{z-scores}, which were interpreted as indicating conventional thought if positive and novel thought if negative. Consequently, every article was associated with multiple z-scores corresponding to co-cited journal pairs in its refernces. For each article, positional statistics of z-scores were calculated to set thresholds for a binary (high/low) classification of conventionality (C) using the median and novelty (N) using the tenth percentile of z-scores within an article. Thus, HCLN would denote high conventionality and low novelty with all four combinations being possible. 
\cite{uzzi_atypical_2013} observed that HCHN articles were twice as likely to be highly cited, suggesting that novel combinations of ideas flavoring a body of conventional thought were a 'nearly universal` feature for impact.

Key to Uzzi {\em et al.}, is their random graph model and its underlying assumptions. The citation switching algorithm used to generate a null model elegantly preserves the number of publications, the number of references in each publication, the year of publication of both publications and references, and the disciplinary composition of the references cited in these publications. However, random substitution of the references used to generate expected values is equiprobable with respect to disciplinary origin and citation count. The model permits substitution of a reference in quantum physics with equal probability by a reference in the related quantum chemistry field or in some entirely different field, such as classical literature, evolutionary biology, or anthropology. Such substitutions poorly model the disciplinary nature of scientific endeavor and citation behavior~\citep{wallace_lariviere_gingras_2012,moed_measuring_2010,klavans_research_2017,garfield_1979}. In addition, under this random model, a reference cited over 100 times in a given year is selected with the same probability as a reference cited only once, which appears inconsistent with the power law or lognormal citation distributions described in the literature~\citep{stringer_statistical_2010,perline_strong_2005}.  Accordingly, model mis-specification is likely to arise from the expected value calculations used to generate z-scores and characterize articles in terms of conventionality and novelty when disciplinary subsets of the WoS are analyzed in the context of expected values generated across all of WoS.

A follow-up study by \citep{boyack_vs_uzzi_2014} explored the impact of discipline and journal effects on the definition of conventionality and novelty.  While their study had some methodological differences in the use of Scopus data rather than WoS data, a smaller dataset, and a $\chi^2$ calculation rather than Monte Carlo simulations to generate expected values of journal pairs, Boyack and Klavans noted strong effects from both disciplines and journals that were not reported by Uzzi et al. While they reported the trend that HCHN is more probable in highly cited papers, they observed that ``only 64.4\%  of  243  WoS  subject  categories'' in the Uzzi {\em et al.} study met the criterion of having the highest probability of hit papers in the HCHN category.  Further, they observed that journals vary widely in terms of size and influence and that 20 journals accounted for nearly 15\% of co-citations in their measurements. Lastly, they noted that three multidisciplinary journals accounted for 9.4\% of all atypical combinations.

Despite different methods used to generate expected values, both of these key preceding studies (\emph{vide supra}) measured co-citation frequencies across the scientific literature and normalized them without disciplinary constraints before subsequently analyzing disciplinary subsets. We hypothesized that normalization using expected values drawn from datasets focused in disciplinary areas, rather than all of WoS, would reduce model mis-specification.  Consequently, we used keyword searches of the scientific literature to construct citation networks themed around academic disciplines: \emph{applied physics, immunology, and metabolism}. Within these disciplinary frameworks, we calculated observed and expected co-citation frequencies using a refined random graph model and an efficient Monte Carlo simulation algorithm.

@@

\section{Materials and Methods}
\emph{Bibliographic data} We have previously developed ERNIE, an open source knowledge platform into which we parse the Web of Science (WoS) Core Collection~\citep{Keserci371955}. WoS data stored in ERNIE spans the period 1900-2019 and consists of over 72 million publications. For this study, we generated an analytical dataset from years 1985 to 2005 from ERNIE. The total number of publications in this dataset was just over 25 million publications (25,134,073), which were then stratified by year of publication. For each of these years, we further restricted analysis to publications of type Article. Since WoS data also contains incomplete references or references that point at other indexes, we also considered only those references for which there were complete records~(Table \ref{tab:summary_data}). For example, WoS data for year 2005 contained 1,753,174 publications, which after restricting to type Article and considering only those references described above resulted in 916,573 publications, 6,095,594 unique references (set of references), and 17,167,347 total references (multiset of references). Given consistent trends in the data, Table \ref{tab:summary_data}, we analyzed the two boundary years (1985 and 2005) and the mid-point (1995). We also used the number of times each of these articles was cited in the first 8 years since publication as a measure of its impact.

% latex table generated in R 3.5.3 by xtable 1.8-4 package
% Fri May 17 14:28:31 2019
\begin{table}[ht]
\caption{Summary of base WoS Analytical Dataset.The number of publications of type Article (UP), unique references (UR), total references (TR) and the ratio of total references to unique references increases monotonically with each year indicating that both the number of documents and citation activity increase over time. Only publications of type Article with at least two references, and references with complete publication data were counted.} 
\label{tab:summary_data}
\centering
\begin{tabular}{|r rrrr|}
  \hline
year & UP & UR & TR & TR/UR \\ 
  \hline
1985 & 391,860 & 2,266,584 & 5,588,861 & 2.47 \\ 
  \hline
1986 & 402,309 & 2,316,451 & 5,708,796 & 2.46 \\ 
1987 & 412,936 & 2,427,347 & 5,998,513 & 2.47 \\ 
1988 & 426,001 & 2,545,647 & 6,354,917 & 2.50 \\ 
1989 & 443,144 & 2,673,092 & 6,749,319 & 2.52 \\ 
1990 & 458,768 & 2,827,517 & 7,209,413 & 2.55 \\ 
1991 & 477,712 & 2,977,784 & 7,729,776 & 2.60 \\ 
1992 & 492,181 & 3,134,109 & 8,188,940 & 2.61 \\ 
1993 & 504,488 & 3,278,102 & 8,676,583 & 2.65 \\ 
1994 & 523,660 & 3,458,072 & 9,255,748 & 2.68 \\ 
  \hline
1995 & 537,160 & 3,680,616 & 9,875,421 & 2.68 \\ 
  \hline
1996 & 663,110 & 4,144,581 & 11,641,286 & 2.81 \\ 
1997 & 677,077 & 4,340,733 & 12,135,104 & 2.80 \\ 
1998 & 693,531 & 4,573,584 & 12,728,629 & 2.78 \\ 
1999 & 709,827 & 4,784,024 & 13,280,828 & 2.78 \\ 
2000 & 721,926 & 5,008,842 & 13,810,746 & 2.76 \\ 
2001 & 727,816 & 5,203,078 & 14,261,189 & 2.74 \\ 
2002 & 747,287 & 5,464,045 & 15,001,390 & 2.75 \\ 
2003 & 786,284 & 5,773,756 & 16,024,652 & 2.78 \\ 
 2004 & 826,834 & 6,095,594 & 17,167,347 & 2.82 \\ 
   \hline
 2005 & 886,648 & 6,615,824 & 19,036,324 & 2.88 \\ 
 \hline
\end{tabular}
\end{table}

We constructed three disciplinary datasets based on keyword searches. (i) immunology (ii) metabolism (iii) applied physics. For the first two, rooted in biomedical research, we searched Pubmed for the term `immunology' or `metabolism' in the years 1985, 1995, and 2005~(Table \ref{tab:disc-datasets}). Pubmed IDs (pmids) returned were matched to WoS IDs (wos\_ids) and used to retrieve relevant articles. For the applied physics dataset, we directly searched subject labels in WoS for `applied physics'. We also examined publications in the five major research areas in the Web of Science; life sciences \& biomedicine, physical sciences, technology, social sciences, arts \& humanities using the extended subcategory classification of 153 sub-groups to categorize disciplinary composition of cited references in the datasets we studied.

\begin{table}[ht]
\caption{Disciplinary Datasets. PubMed and WoS were searched for articles using search terms, `immunology', `metabolism', and `applied physics'. Counts of publications are shown for each of the three years analyzed.}
\label{tab:disc-datasets}
\centering
\begin{tabular}{|r r r r|}
  \hline
Year & Immunology & Metabolism & Applied Physics \\ 
  \hline
1985 & 21,606 & 78,998 & 10,298 \\ 
1995 & 29,320 & 121,247 & 21,012  \\ 
2005 & 37,296 & 200,052 & 35,600  \\ 
 \hline
\end{tabular}
\end{table}

@@

We developed a performant citation switching algorithm, \emph{runtime enhanced permuting citation switcher (repcs)} that randomly permuted citations grouped by year of publication to switch citations while preserving the year of publication for both articles and references, the number of publications and the number of references in each dataset and the disciplinary composition of the references in each dataset. We used the citation switching algorithm of ~\cite{uzzi_atypical_2013} referred to as \emph{umsj}~\citep{boyack_vs_uzzi_2014}, as a benchmark for the repcs algorithm using code kindly provided by the authors. A comparative analysis showed that while 10 simulations of the WoS 1985 dataset (391,860 publications) completed in 2,186 hours using the umsj algorithm, it completed in less than one hour using the repcs algorithm implemented on a Spark cluster. Using either the repcs or umsj algorithms for 10 simulations resulted in expected value coverage of 75\% of the observed journal pair frequencies. Subsequent rounds of development and testing resulted in 99\% coverage of observed journal pairs when 1,000 simulations were conducted. The runtime advantage was significant enough that we chose to use the repcs algorithm in our study  and generated expected values averaged from 1,000 simulations for improved coverage for every dataset we analyzed.

Averaging the result of 1,000 simulations for each dataset studied, z-scores were then calculated for each journal-pair using the formula $(F_{obs} - F_{exp})/\sigma$ where $F_{obs}$ is the observed frequency, $F_{exp}$ is the averaged simulated frequency, and $\sigma$ is the standard deviation of the simulated frequencies. As a result of these calculations, each publication becomes associated with a set of z-scores corresponding to the journal pairs derived from pairwise combinations of its cited references. Positional statistics of z-scores were calculated for each publication, which was then labeled according to conventionality and novelty: (i) HC if the median z-score exceeded the median of median z-scores for all publications and LC if the median z-score was equal to or less than the median of median z-scores for all publications, and (ii) HN if the tenth percentile of z-scores for a publication was less than zero, and LN if the tenth percentile of z-scores for a publication was greater than zero.

@@

\subsection{Model Misspecification and the Attributes of Disciplinary Context} We invoke model mis-specification arising from irrelevant references as a concern when disciplinary subsets are analyzed using a random graph model that generates expected co-citation frequencies from the full body of scientific literature. A symptom of model misspecification is that random sampling of citations associated with an article do not match with the distribution of citations across the disciplines observed in articles from that discipline. \cite{wallace_lariviere_gingras_2012} found that, often, a majority of an article's citations are from the specialty of the article, although that percentage varied among disciplines in the eight specialties they investigated: from approximately 39\% to 89\% for 2006. To address this issue, we first assembled three exemplar datasets using keyword searches for the terms `applied physics', `immunology', and `metabolism' (Materials and Methods). Using the WoS extended classificatation of 153 subject areas, we then analyzed the disciplinary composition of the references in these three datasets into the five research areas in the Web of Science (life sciences and biomedicine, physical sciences, social sciences, technology, and arts and humanities) for the years 1985, 1995, and 2005. We observed that 84.2\% of the references in the applied physics dataset were classified as belonging to the WoS physical sciences research area. Similarly, 93.7\% and 92.1\% of the references in the immunology and metabolism datasets belonged to the life sciences and biomedicine research area, indicating strongly that references are more likely belong to the same research area as the publications that cite them.

 We also analyzed the consequences of citation shuffling within a disciplinary set or all of the Web of Science.  References in publications belonging to these three datasets were classified using the WoS Extended Subjects classification of 153 subject areas and summarized as a frequency distribution. A single shuffle of the references in these datasets or all the references in the corresponding WoS year slice was performed, using either the rpcs or umsj algorithms, after which subject frequencies were computed again. The fold change in subject frequencies of references before and after shuffling was calculated for these groups using all 153 Subject categories and are summarized as boxplots in Fig \ref{fig:be}. For example, the applied physics dataset contained 1 reference labeled Genetics and Heredity . This proportion was preserved in the intra-network shuffle but increased 1496 times when the WoS background was used as a source of substitution during Monte Carlo simulation with the repcs algorithm. Similarly, with the metabolism dataset.which contained one reference labeled Philosophy that was increased to 661 after a single repcs shuffle. The data show convincingly that the disciplinary composition of references in a network is preserved when citation shuffling is conducted within it but is significantly altered when the full WoS network is used as a source of substitution. A second inference is that the choice between these two algorithms is not relevant beyond runtime advantage but that the background network used for the shuffle is critical.

@@

\begin{figure}%[tbhp]
\centering
\includegraphics[width=0.8\linewidth]{background-effect}     
\caption{\ Intra-network citation shuffling preserves the disciplinary composition of references. Publications of type Article belonging to the three disciplinary networks (i) Applied Physics (ii) Immunology (iii) Metabolism were subject to a single shuffle of all their cited references using either the cited references in these networks as a source of random substitutions (bg\_local) or references from all articles in WoS (bg\_allWoS). Citation shuffling was performed using either our algorithm (repcs) or that of Uzzi (umsj). The disciplinary composition of cited references before and after shuffling was measured as frequencies for each of 153 sub-disciplines (from the extended subject classification in WoS) and expressed as fold difference between citation counts grouped by subject for original (o) and shuffled (s) references using the formula (fold\_difference = $ifelse(o > s, o/s, s/o)$) and rounded to the nearest integer. A fold difference of 1 indicates that citation shuffling did not alter disciplinary composition. Data are shown for articles published in 1985. All eight boxplots are generated from 153 observations. Null values were set to 1. Note y-axis: $log_2$ scale.} 
\label{fig:be}
\end{figure}

@@

\authorcontributions 
This study was designed by GC, JB, SD, and TW. Simulations and analysis were performed by AD, GC, JB, and SD. Infrastructure and workflows used to generate data used in this study were developed by AD, DK, SL, SD, and GC.  All authors reviewed and commented on the manuscript, which was written by GC, JB, and TW.

>>>>

%% Documentclass:
%% Network Neuroscience
%\documentclass[finalfonts,NETN]{stjour}
\documentclass[NETN]{stjour}
\usepackage{comment}

@@

% CREATED A BRANCH FOR SAFETY %
%%%%%%%%%%% Article Set-Up %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Article Type:
%% Default is Research.

@@

Citation and network analysis of  scientific literature reveals information on semantic relationships between publications, collaboration between scientists, and the practice of citation itself~\citep{garfield_citation_1955,de_solla_price_networks_1965,newman_structure_2001,Shi:2010:CHI:1816123.1816131,patience_pmid28560354}. Co-citation, the frequency with which two documents are cited together in other documents provides additional insights, including the identification of semantically related documents, fields, specializations, and new ideas in science~\citep{small_co-citation_1973, marshakova-shaikevich_co-citation_1973,boyack_co-citation_2010, 10.3389/frma.2018.00020}.

\cite{uzzi_atypical_2013} used a novel approach for co-citation analysis to characterize a subset of highly cited articles with respect to both novel and conventional combinations of prior research. The frequency with which references were co-cited  in 17.9 million articles and their cited references from the Web of Science (WoS) was calculated and expressed as journal pair frequencies (observed co-citation frequencies). Expected co-citation values were generated from Monte Carlo simulations under a random graph model. Observed frequencies were then normalized (shifted and scaled) to averaged expected values from ten simulations and termed as \emph{z-scores}. Consequently, every article was associated with multiple z-scores corresponding to co-cited journal pairs in its references. For each article, positional statistics of z-scores were calculated to set thresholds for a binary (high/low) classification of conventionality (C) using the median z-score of an article and novelty (N) using the tenth percentile of z-scores within an article. Thus, HCLN would denote high conventionality and low novelty with all four combinations being possible. The authors observed that HCHN articles were twice as likely to be highly cited, suggesting that novel combinations of ideas flavoring a body of conventional thought were a feature of impact and that `science follows a nearly universal pattern'.

Key to Uzzi {\em et al.}, is their random graph model and its underlying assumptions. The citation switching algorithm used to generate expected values from random substitutions is designed to preserve the number of publications, the number of references in each publication, the year of publication of both publications and references, and the disciplinary composition of the references cited in these publications. However, substitutions are equiprobable with respect to disciplinary origin and citation count. The model permits substitution of a reference in quantum physics with equal probability by a reference in quantum physics, quantum chemistry, classical literature, entomology, or anthropology. Such substitutions poorly model the disciplinary nature of scientific endeavor and citation behavior~\citep{wallace_lariviere_gingras_2012,moed_measuring_2010,klavans_research_2017,garfield_1979}. In addition, under this random model, a reference cited over 100 times in a given year is selected with the same probability as a reference cited only once, which appears inconsistent with the power law or lognormal citation distributions described in the literature~\citep{stringer_statistical_2010,perline_strong_2005}.  Accordingly, model mis-specification is likely to arise on account of the simulated values not modeling the empirical data very well.

A follow-up study by \cite{boyack_vs_uzzi_2014} explored the impact of discipline and journal effects on these definition of conventionality and novelty.  While their study had some methodological differences in the use of Scopus data rather than WoS data, a smaller dataset, and a $\chi^2$ calculation rather than Monte Carlo simulations to generate expected values of journal pairs, Boyack and Klavans noted strong effects from both disciplines and journals that were not reported by Uzzi et al. While they reported the trend that HCHN is more probable in highly cited papers, they observed that ``only 64.4\%  of  243  WoS  subject  categories'' in the Uzzi {\em et al.} study met the criterion of having the highest probability of hit papers in the HCHN category.  Further, they observed that journals vary widely in terms of size and influence and that 20 journals accounted for nearly 15\% of co-citations in their measurements. Lastly, they noted that three multidisciplinary journals accounted for 9.4\% of all atypical combinations.

Despite different methods used to generate expected values, both of these key preceding studies (\emph{vide supra}) measured co-citation frequencies across the scientific literature (the WoS superset) and normalized them without disciplinary constraints before subsequently analyzing disciplinary subsets. We hypothesized that normalization using expected values drawn from datasets focused in disciplinary areas, rather than all of WoS, would reduce model mis-specification.  Consequently, we used keyword searches of the scientific literature to construct exemplar citation networks themed around academic disciplines: \emph{applied physics, immunology, and metabolism}. Within these disciplinary frameworks, we calculated observed and expected co-citation frequencies using a refined random graph model and an efficient Monte Carlo simulation algorithm.

@@

\section{Materials and Methods}
\emph{Bibliographic data} We have previously developed ERNIE, an open source knowledge platform into which we parse the Web of Science (WoS) Core Collection~\citep{Keserci371955}. WoS data stored in ERNIE spans the period 1900-2019 and consists of over 72 million publications. For this study, we generated an analytical dataset from years 1985 to 2005 using data in ERNIE. The total number of publications in this dataset was just over 25 million publications (25,134,073), which were then stratified by year of publication. For each of these years, we further restricted analysis to publications of type Article. Since WoS data also contains incomplete references or references that point at other indexes, we also considered only those references for which there were complete records~(Table \ref{tab:summary_data}). For example, WoS data for year 2005 contained 1,753,174 publications, which after restricting to type Article and considering only those references described above resulted in 916,573 publications, 6,095,594 unique references (set of references), and 17,167,347 total references (multiset of references). Given consistent trends in the data, Table \ref{tab:summary_data}, we analyzed the two boundary years (1985 and 2005) and the mid-point (1995). We also used the number of times each of these articles was cited in the first 8 years since publication as a measure of its impact.

% latex table generated in R 3.5.3 by xtable 1.8-4 package
% Fri May 17 14:28:31 2019
\begin{table}[ht]
\caption{Summary of base WoS Analytical Dataset.The number of publications of type Article (UP), unique references (UR), total references (TR) and the ratio of total references to unique references increases monotonically with each year indicating that both the number of documents and citation activity increase over time. Only publications of type Article with at least two references, and references with complete publication data were included.} 
\label{tab:summary_data}
\centering
\begin{tabular}{|r rrrr|}
  \hline
year & UP & UR & TR & TR/UR \\ 
  \hline
1985 & 391,860 & 2,266,584 & 5,588,861 & 2.47 \\ 
  \hline
1986 & 402,309 & 2,316,451 & 5,708,796 & 2.46 \\ 
1987 & 412,936 & 2,427,347 & 5,998,513 & 2.47 \\ 
1988 & 426,001 & 2,545,647 & 6,354,917 & 2.50 \\ 
1989 & 443,144 & 2,673,092 & 6,749,319 & 2.52 \\ 
1990 & 458,768 & 2,827,517 & 7,209,413 & 2.55 \\ 
1991 & 477,712 & 2,977,784 & 7,729,776 & 2.60 \\ 
1992 & 492,181 & 3,134,109 & 8,188,940 & 2.61 \\ 
1993 & 504,488 & 3,278,102 & 8,676,583 & 2.65 \\ 
1994 & 523,660 & 3,458,072 & 9,255,748 & 2.68 \\ 
  \hline
1995 & 537,160 & 3,680,616 & 9,875,421 & 2.68 \\ 
  \hline
1996 & 663,110 & 4,144,581 & 11,641,286 & 2.81 \\ 
1997 & 677,077 & 4,340,733 & 12,135,104 & 2.80 \\ 
1998 & 693,531 & 4,573,584 & 12,728,629 & 2.78 \\ 
1999 & 709,827 & 4,784,024 & 13,280,828 & 2.78 \\ 
2000 & 721,926 & 5,008,842 & 13,810,746 & 2.76 \\ 
2001 & 727,816 & 5,203,078 & 14,261,189 & 2.74 \\ 
2002 & 747,287 & 5,464,045 & 15,001,390 & 2.75 \\ 
2003 & 786,284 & 5,773,756 & 16,024,652 & 2.78 \\ 
 2004 & 826,834 & 6,095,594 & 17,167,347 & 2.82 \\ 
   \hline
 2005 & 886,648 & 6,615,824 & 19,036,324 & 2.88 \\ 
 \hline
\end{tabular}
\end{table}

We constructed three disciplinary datasets based on keyword searches. (i) immunology (ii) metabolism (iii) applied physics. For the first two, rooted in biomedical research, we searched Pubmed for the term `immunology' or `metabolism' in the years 1985, 1995, and 2005~(Table \ref{tab:disc-datasets}). Pubmed IDs (pmids) returned were matched to WoS IDs (wos\_ids) and used to retrieve relevant articles. For the applied physics dataset, we directly searched traditional subject labels in WoS for `applied physics'. We also examined publications in the five major research areas in the Web of Science; life sciences \& biomedicine, physical sciences, technology, social sciences, arts \& humanities using the extended subcategory classification of 153 sub-groups to categorize disciplinary composition of cited references in the datasets we studied.

\begin{table}[ht]
\caption{Disciplinary Datasets. PubMed and WoS were searched for articles using search terms, `immunology', `metabolism', and `applied physics'. Counts of publications are shown for each of the three years analyzed and expressed in parentheses as a percentage of the total number of publications in our analytical WoS dataset (Table \ref{tab:summary_data}) for that year.}
\label{tab:disc-datasets}
\centering
\begin{tabular}{|r r r r|}
  \hline
Year & Immunology & Metabolism & Applied Physics \\ 
  \hline
1985 & 21,606 (5.5\%) & 78,998 (20.2\%) & 10,298 (2.7\%) \\ 
1995 & 29,320 (5.5\%)  & 121,247 (22.6\%) & 21,012 (3.9\%)  \\ 
2005 & 37,296 (4.2\%) & 200,052 (22.6\%) & 35,600 (4.0\%)  \\ 
 \hline
\end{tabular}
\end{table}

@@

For citation shuffling, we developed a performant citation switching algorithm, \emph{runtime enhanced permuting citation switcher (repcs)} that randomly permuted citations grouped by year of publication to switch citations while preserving the year of publication for both articles and references, the number of publications and the number of references in each dataset and the disciplinary composition of the references in each dataset. In implementing this approach, our approach differs from umsj in the following ways (i) we sampled citations from a multiset rather than a set in order to reflect the observed frequency of citation (ii) we permitted a substitution to match the original reference in a publication when the random selection process dictated it rather than attempting to enforce that a different reference be substituted (iii) we introduced an error correction step to delete any publications that accumulated duplicate references during the substitution process. As a benchmark, we used the citation switching algorithm of ~\cite{uzzi_atypical_2013} referred to by others as \emph{umsj}~\citep{boyack_vs_uzzi_2014}, using code kindly provided by the authors. A single comparative analysis showed that while 10 simulations of the WoS 1985 dataset (391,860 publications) completed in 2,186 hours using the umsj algorithm, it completed in less than one hour using the our implementation of the repcs algorithm on a Spark cluster. We also tested repcs under comparable conditions to umsj and estimated a runtime advantage of at least two orders of magnitude. Using either the repcs or umsj algorithms for 10 simulations resulted in expected value coverage of 75\% of the observed journal pair frequencies. Subsequent rounds of development and testing resulted in 99\% coverage of observed journal pairs when 1,000 simulations were conducted. The runtime advantage was significant enough that we chose to use the repcs algorithm in our study  and generated expected values averaged from 1,000 simulations for improved coverage of every dataset we analyzed.

Averaging the result of 1,000 simulations for each dataset studied, z-scores were then calculated for each journal-pair using the formula $(F_{obs} - F_{exp})/\sigma$ where $F_{obs}$ is the observed frequency, $F_{exp}$ is the averaged simulated frequency, and $\sigma$ is the standard deviation of the simulated frequencies for a journal pair. As a result of these calculations, each publication becomes associated with a set of z-scores corresponding to the journal pairs derived from pairwise combinations of its cited references. Positional statistics of z-scores were calculated for each publication, which was then labeled according to conventionality and novelty: (i) HC if the median z-score exceeded the median of median z-scores for all publications and LC if the median z-score was equal to or less than the median of median z-scores for all publications, and (ii) HN if the tenth percentile of z-scores for a publication was less than zero, and LN if the tenth percentile of z-scores for a publication was greater than zero.

@@

\subsection{Model Misspecification and the Attributes of Disciplinary Context} \emph{This is where we want to add the additional data about expected values ~2.5\% for AP as an example.} We invoke model mis-specification due to considering disciplinarily irrelevant references as a concern when disciplinary subsets are analyzed using a random graph model that generates expected co-citation frequencies from the WoS superset. A symptom of model misspecification is that random sampling of citations associated with an article do not match with the distribution of citations across the disciplines observed in articles from that discipline. To address this issue, we first assembled three exemplar datasets using keyword searches for the terms `applied physics', `immunology', and `metabolism' (Materials and Methods). Using the WoS extended classificatation of 153 subject areas, we then analyzed the disciplinary composition of the references in these three datasets into the five research areas in the Web of Science (life sciences and biomedicine, physical sciences, social sciences, technology, and arts and humanities) for the years 1985, 1995, and 2005. We observed that 84.2\% of the references in the applied physics dataset were classified as belonging to the WoS physical sciences research area. Similarly, 93.7\% and 92.1\% of the references in the immunology and metabolism datasets belonged to the life sciences and biomedicine research area, indicating strongly that references are more likely belong to the same research area as the publications that cite them. Our observations are in agreement with \cite{wallace_lariviere_gingras_2012} who found that, often, a majority of an article's citations are from the specialty of the article, although that percentage varied among disciplines in the eight specialties they investigated: from approximately 39\% to 89\% for 2006.

We also analyzed the consequences of citation shuffling within a disciplinary set or all of the Web of Science.  References in publications belonging to these three datasets were classified using the WoS Extended Subjects classification of 153 subject areas and summarized as a frequency distribution. A single shuffle of the references in these datasets or all the references in the corresponding WoS year slice was performed, using either the rpcs or umsj algorithms, after which subject frequencies were computed again. The fold change in subject frequencies of references before and after shuffling was calculated for these groups using all 153 Subject categories and are summarized as boxplots in Fig \ref{fig:be}. For example, the applied physics dataset contained 1 reference labeled Genetics and Heredity . This proportion was preserved in the intra-network shuffle but increased 1496 times when the WoS background was used as a source of substitution during Monte Carlo simulation with the repcs algorithm. Similarly, with the metabolism dataset.which contained one reference labeled Philosophy that was increased to 661 after a single repcs shuffle. The data show convincingly that the disciplinary composition of references in a network is preserved when citation shuffling is conducted within it but is significantly distorted when the WoS superset is used as a source of substitution. A second inference is that the choice between these two algorithms is not relevant beyond runtime advantage but that the background network used for the shuffle is critical.

@@

\begin{figure}%[tbhp]
\centering
\includegraphics[width=0.6\linewidth]{background-effect}     
\caption{\ Intra-network citation shuffling preserves the disciplinary composition of references. Publications of type Article belonging to the three disciplinary networks (i) Applied Physics (ii) Immunology (iii) Metabolism were subject to a single shuffle of all their cited references using either the cited references in these networks as a source of random substitutions (bg\_local) or references from all articles in WoS (bg\_allWoS). Citation shuffling was performed using either our algorithm (repcs) or that of Uzzi (umsj). The disciplinary composition of cited references before and after shuffling was measured as frequencies for each of 153 sub-disciplines (from the extended subject classification in WoS) and expressed as fold difference between citation counts grouped by subject for original (o) and shuffled (s) references using the formula (fold\_difference = $ifelse(o > s, o/s, s/o)$) and rounded to the nearest integer. A fold difference of 1 indicates that citation shuffling did not alter disciplinary composition. Data are shown for articles published in 1985. All eight boxplots are generated from 153 observations each. Null values were set to 1. Note y-axis: $log_2$ scale.} 
\label{fig:be}
\end{figure}

@@

\authorcontributions 
This study was designed by GC, JB, SD, and TW. Simulations and analysis were performed by AD, DLH, GC, JB, and SD. Infrastructure and workflows used to generate data used in this study were developed by AD, DK, SL, SD, and GC.  All authors reviewed and commented on the manuscript, which was written by GC, JB, and TW.
